---
title: "Documentation for analysis.R — Monthly Sales Analysis (Synthetic Data)"
author: "Paul"
date: "2026-08-19"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
    highlight: tango
---

> **AI Assistance Declaration**: I used Claude (Anthropic, Claude Fable 5, 2026-08-19) for Markdown formatting suggestions and section organization of this document. Prompts used: "Here's my R script description. Suggest Markdown formatting and code block examples.", "Add syntax highlighting and improve section organization (e.g., # Purpose, ## Inputs, ## Outputs).", and "Is the Markdown consistent with RMarkdown best practices?". I verified outputs by knitting this file and checking that every chunk runs without errors. All final calculations are done by myself. I am responsible for the accuracy and originality of this work.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

`analysis.R` performs a monthly sales analysis on a **synthetic** retail dataset. It loads the data, checks it for missing values, engineers a `revenue` variable, computes summary statistics, aggregates revenue by month and by product category, and saves the results (two CSV files and one chart) to an `output/` folder.

The script is written in **base R only**, so it runs on any standard R installation with no extra packages.

## Inputs

The script expects a single input file, relative to the project root:

| Input                      | Format | Description                                        |
|----------------------------|--------|----------------------------------------------------|
| `data/synthetic_sales.csv` | CSV    | 300 synthetic transactions, Jan–Jun 2026, 7 columns |

The dataset columns are `transaction_id`, `date`, `category`, `product_id`, `units_sold`, `unit_price`, and `region`. It was generated with `set.seed(400)`, so it is fully reproducible and contains **no real or personal data**.

## Outputs

| Output                            | Description                                 |
|-----------------------------------|---------------------------------------------|
| `output/monthly_revenue.csv`      | Total revenue per month                     |
| `output/category_summary.csv`     | Revenue and units sold per product category |
| `output/monthly_revenue_plot.png` | Bar chart of monthly revenue                |

The script also prints row counts, a structure summary, and a table of summary statistics to the console.

# How the Script Works

## Step 1 — Load the data

The CSV is read with `read.csv()` and the `date` column is converted from text to R's `Date` class so it can be formatted by month later.

```{r load-data}
sales <- read.csv("data/synthetic_sales.csv", stringsAsFactors = FALSE)
sales$date <- as.Date(sales$date)
head(sales)
```

## Step 2 — Basic data checks

Before any calculation, the script confirms how many rows were loaded and that there are no missing values.

```{r data-checks}
cat("Rows loaded:", nrow(sales), "\n")
cat("Missing values:", sum(is.na(sales)), "\n")
```

## Step 3 — Feature engineering

Two derived columns are added: `revenue` (units × price) and `month` (a `YYYY-MM` key used for aggregation).

```{r feature-engineering}
sales$revenue <- sales$units_sold * sales$unit_price
sales$month   <- format(sales$date, "%Y-%m")
```

## Step 4 — Summary statistics

A small data frame collects the headline numbers: total revenue, mean revenue per sale, median units sold, and the transaction count.

```{r summary-stats}
summary_stats <- data.frame(
  metric = c("Total revenue", "Mean revenue per sale",
             "Median units sold", "Number of transactions"),
  value  = c(round(sum(sales$revenue), 2),
             round(mean(sales$revenue), 2),
             median(sales$units_sold),
             nrow(sales))
)
knitr::kable(summary_stats, caption = "Summary statistics (synthetic data)")
```

## Step 5 — Aggregations

`aggregate()` groups revenue by month, and revenue plus units sold by category. Categories are sorted by revenue, highest first.

```{r aggregations}
monthly_revenue <- aggregate(revenue ~ month, data = sales, FUN = sum)
monthly_revenue$revenue <- round(monthly_revenue$revenue, 2)

category_summary <- aggregate(
  cbind(revenue, units_sold) ~ category, data = sales, FUN = sum
)
category_summary$revenue <- round(category_summary$revenue, 2)
category_summary <- category_summary[order(-category_summary$revenue), ]

knitr::kable(category_summary, row.names = FALSE,
             caption = "Revenue and units sold by category")
```

## Step 6 — Save outputs and plot

The results are written to the `output/` folder, and a bar chart of monthly revenue is drawn with base R's `barplot()`.

```{r plot, fig.width=8, fig.height=5, fig.cap="Monthly revenue from the synthetic dataset"}
barplot(monthly_revenue$revenue,
        names.arg = monthly_revenue$month,
        col       = "#4C72B0",
        las       = 2,
        main      = "Monthly Revenue (Synthetic Data)",
        ylab      = "Revenue ($)")
```

In `analysis.R`, this chart is saved to `output/monthly_revenue_plot.png` via `png()` and `dev.off()`, and the two summary tables are saved with `write.csv()`.

# How to Run

From the project root, either source the script inside R:

```r
source("analysis.R")
```

or run it from a terminal:

```bash
Rscript analysis.R
```

# Session Info

```{r session-info}
sessionInfo()
```
